Estimating the genome-wide contribution of selection to temporal allele frequency change

### 2020-05-05 manuscript/manuscript.tex
more Gs changes

\begin{align}
  \vb{G(t) = \frac{ \sum_{i=0}^{t-1} \sum_{j \ne i}^{t-1} \cov(\Delta p_i, \Delta p_j)}{\var(p_t - p_0)}}
\end{align}
%
which estimates the effect of selection on allele frequency change between the
initial generation $0$ and some later generation $t$, which can be varied to
see how this quantity grows through time. Since \textcite{Barghi2019-qy}
experiment is sequenced every ten generations, in the numerator for the
covariance we use the allele frequency changes between adjacent timepoints,
which are ten generations apart; \vb{thus the strong covariances between
  adjacent generations are not observable and do not contribute to the
  numerator of $G(t)$}. \vb{Additionally, selection also inflates the variance
  in allele frequency change per generation; however, this effect cannot be
  easily distinguished from drift.} \vb{For both these reasons, our measure
  $G(t)$ is strongly conservative} (\vb{we demonstrate this through simulations
in Supplementary Material Section \ref{supp:tempblocks}}). Still, we find a
remarkably strong signal. Greater than $20\%$ of total, genome-wide allele
frequency change over 60 generations is the result of selection (Figure
\ref{fig:figure-1} B).

@@

  \caption{{\bf A}: The convergence \vb{correlations}, averaged across
    \textcite{Barghi2019-qy} replicate pairs, through time. Each line
    represents the convergence correlation $\mathrm{cor}(\Delta p_{s}, \Delta
    p_{s})$ from a starting reference generation $s$ to a later time $t$, which
    varies along the x-axis; each line corresponds to a row of the temporal
    convergence correlation matrix depicted to the right \vb{(where the
      diagonal elements represent the convergence correlations between the same
      timepoints across replicate populations)}. We note that convergent
      correlation for the last timepoint is an outlier; we are unsure as to the
      cause of this, e.g. it does not appear to be driven by a single pair of
      replicates. {\bf B}: The convergence \vb{correlations} between individual
      pairs of replicates in the \textcite{Kelly2019-dc} data (\vb{note the
      confidence intervals are plotted, but are small on this y-axis scale}).
      {\bf C}:  The convergence \vb{correlations} between individual pairs of
      replicates in \parencite{Castro2019-uk} data, for the two selection lines
      (LS1 and LS2) and the control (Ctrl); gray CIs are those using the
      complete dataset, blue CIs exclude chromosomes 5 and 10 which harbor the
      two regions \textcite{Castro2019-uk} found to have signals of parallel
      selection between LS1 and LS2. \vb{Through simulations, we have found
        that the differences in convergence correlation confidence interval
        widths between these \emph{Drosophila} studies and the Longshanks study
        are due to the differing population sizes.}}

@@

Finally, we observed that in the longest study  we analyzed
\parencite{Barghi2019-qy}, some genome-wide temporal covariances become
negative at future timepoints (see the first two rows in Figure
\ref{fig:figure-1} A). This shows that alleles that were on average going up
initially are later going down in frequency, i.e. that the average direction of
selection experienced by alleles has flipped. This must reflect either a change
in the environment or the genetic background, due to epistatic relationships
among alleles altered by frequency changes or recombination breaking up
selective alleles.  Such reversals of selective dynamics could be occurring at
other timepoints but the signal of a change in the direction of selection at
particular loci may be washed out when we calculate our genome-wide average
temporal covariances. To address this limitation, we calculated the
distribution of the temporal covariances over 100kb windowed covariances
(Figure \ref{fig:figure-3} shows these distributions pooling across all
replicates; see Supplementary Figure
\ref{suppfig:barghi-offset-replicate-panels} for individuals replicates). The
covariance estimate of each genomic window will be noisy, due to sampling and
genetic drift, and the neutral distribution of the covariance is complicated
due to linkage disequilibria (which can occur over long physical distances in
E\&R and selection studies, \cite{Nuzhdin2013-gf,Baldwin-Brown2014-cl}). To
address this, we have developed a permutation-based procedure that constructs
an empirical null distribution by randomly flipping the signs of the allele
frequency changes per-genomic window. This destroys the systematic covariances
created by linked selection and creates a sampling distribution of the
covariances spuriously created by neutral genetic drift while preserving the
complex dependencies between adjacent loci created by linkage disequilibrium.
This empirical neutral null distribution is conservative in the sense that the
variances of the covariances are wider than expected under drift alone as they
include the effect of selection on the allele frequency change within a
time-interval, just not between time-intervals. We see (Figure
\ref{fig:figure-3} A and B) that windowed temporal covariances between close
timepoints are skewed positive (a heavy right tail), while between more distant
timepoints these windowed temporal covariances tend to shift to become more
negative (a heavy left tail).  We quantified the degree to which the left and
right tails are inflated compared to the null distribution as a function of
time, and see excesses in both tails in Figure \ref{fig:figure-3} C. This
finding is also robust to sign-permuting allele frequency changes on a
chromosome-level, the longest extent that gametic linkage disequilibria can
extend (Supplementary Figure \ref{suppfig:barghi-tailprobs-seqid}). We see a
striking pattern that the windowed covariances not only decay towards zero, but
in fact become negative through time, consistent with many regions in the
genome having had a reversed fitness effect at later timepoints.

@@

Finally we used forward-in-time simulations to explore the conditions under
which temporal and convergent correlations arise. We show a subset of our
results for a model of Gaussian stabilizing selection (GSS) on a phenotype
where directional selection is induced by a sudden shift in the optimum
phenotype of varying magnitudes (Figure \ref{fig:figure-4}A).  We find that
positive temporal covariances are produced by such selection (Figure
\ref{fig:figure-4}B), and that these positive temporal covariances can compound
together to generate a large proportion of allele frequency change being due to
selection (i.e. large G) over the relatively short time periods similar to our
analyzed selection datasets span (Figure \ref{fig:figure-4}C). The magnitude of
these covariances increases with the strength of selection, i.e. the variance
in fitness, such that stronger selection generates larger proportions of allele
frequency change.  We find a similar picture of stronger convergent selection
pressures generating larger convergence correlations.

These results are often relatively insensitive to the number of loci underlying
the trait, suggesting that they may be capturing highly polygenic signals.
Indeed if only a small number of loci influence the trait, the $G(t)$
trajectories are typically much more stochastic, such that replicates do not
reliably generate temporal covariances (compare Figure \ref{fig:figure-1}B to
Supplementary Material Figure \ref{suppfig:sim-expfit-covs}).

Additionally, we explored other selection schemes. We find that the long term
dynamics of the covariances under truncation selection, which generates
substantial epistasis, are richer than we see under GSS and multiplicative
selection (Supplementary Material Figure \ref{suppfig:supp-trunc}).  We also
conducted simulations of purifying selection alone (i.e. background selection)
and find that this can also generate positive temporal covariances
(Supplementary Material Figure \ref{suppfig:supp-bgs-covs}), thus it is
unlikely that all of the allele frequency change we see is the result of the
novel selection pressure the populations are exposed to, and some of this
selection may be ongoing purifying selection. Only in the case of the
Longshanks experiment, does the control line mean that we can quantify
selection that is due to the novel selection pressure.

While none of our experiments have selected the populations in divergent
directions, in our simulations we find that such selection can generate
negative convergent correlations. This suggests that selection experiments
combining multiple replicates, control lines, as well as divergent selection
pressures might be quite informative for unpacking the contribution of
particular selection pressures to genome-wide allele frequency change.

@@

Our analyses reveal that a sizable proportion of allele frequency change in
these populations is due to the (indirect) action of selection. Capitalizing on
replicated designs, we characterized the extent to which convergent selection
pressures lead to parallel changes in allele frequencies across replicate
populations, and found that a reasonable proportion of the response is shared
across short timescales. These likely represent substantial under-estimates of
the contribution of linked selection because the studies we have reanalyzed do
not sequence the population each generation, preventing us from including the
effects of stronger correlations between adjacent generations.  Furthermore,
our estimation methods are intentionally conservative, for example they exclude
the contribution of selection that does not persist across generations and
selection that reverses sign; thus they can be seen as a strong lower bound of
the effects of selection, \vb{which we have confirmed through forward-in-time
simulations}.

@@

@@

\subsubsection{Sampling in Temporal Blocks}
\label{supp:tempblocks}

>>>>

\begin{align}
  G(t) = \frac{\sum_{i\ne j}^{t-1} \cov(\Delta p_i, \Delta p_j)}{\var(p_t - p_0)}
\end{align}
%
which estimates the effect of selection on allele frequency change between the
initial generation $0$ and some later generation $t$, which can be varied to
see how this quantity grows through time. Since \textcite{Barghi2019-qy}
experiment is sequenced every ten generations, in the numerator for the
covariance we use the allele frequency changes between adjacent timepoints,
which are ten generations apart. Consequently, this leads our measure $G(t)$ to
be strongly conservative, since the temporal covariances within each
ten-generation block are not directly observable, and thus are not included in
the numerator of $G(t)$. Still, we find a remarkably strong signal. Greater
than $20\%$ of total, genome-wide allele frequency change over 60 generations
is the result of selection (Figure \ref{fig:figure-1} B).

@@

  \caption{{\bf A}: The convergence \vb{correlations}, averaged across
    \textcite{Barghi2019-qy} replicate pairs, through time. Each line
    represents the convergence correlation $\mathrm{cor}(\Delta p_{s}, \Delta
    p_{s})$ from a starting reference generation $s$ to a later time $t$, which
    varies along the x-axis; each line corresponds to a row of the temporal
    convergence correlation matrix depicted to the right. We note that
    convergent correlation for the last timepoint is an outlier; we are unsure
    as to the cause of this, e.g. it does not appear to be driven by a single
    pair of replicates. {\bf B}: The convergence \vb{correlations} between
    individual pairs of replicates in the \textcite{Kelly2019-dc} data
    (\vb{note the confidence intervals are plotted, but are small on this
    y-axis scale; see note XXX}). {\bf C}:  The convergence \vb{correlations}
    between individual pairs of replicates in \parencite{Castro2019-uk} data,
    for the two selection lines  (LS1 and LS2) and the control (Ctrl); gray CIs
    are those using the complete dataset, blue CIs exclude chromosomes 5 and 10
    which harbor the two regions \textcite{Castro2019-uk} found to have signals
  of parallel selection between LS1 and LS2.}

@@

%Finally, we

\gc{We} observed that in the longest study  we analyzed
\parencite{Barghi2019-qy}, some genome-wide temporal covariances become
negative at future timepoints (see the first two rows in Figure
\ref{fig:figure-1} A). This shows that alleles that were on average going up
initially are later going down in frequency, i.e. that the average direction of
selection experienced by alleles has flipped. This must reflect either a change
in the environment or the genetic background, due to epistatic relationships
among alleles altered by frequency changes or recombination breaking up
selective alleles.  Such reversals of selective dynamics could be occurring at
other timepoints but the signal of a change in the direction of selection at
particular loci may be washed out when we calculate our genome-wide average
temporal covariances. To address this limitation, we calculated the
distribution of the temporal covariances over 100kb windowed covariances
(Figure \ref{fig:figure-3} shows these distributions pooling across all
replicates; see Supplementary Figure
\ref{suppfig:barghi-offset-replicate-panels} for individuals replicates). The
covariance estimate of each genomic window will be noisy, due to sampling and
genetic drift, and the neutral distribution of the covariance is complicated
due to linkage disequilibria (which can occur over long physical distances in
E\&R and selection studies, \cite{Nuzhdin2013-gf,Baldwin-Brown2014-cl}). To
address this, we have developed a permutation-based procedure that constructs
an empirical null distribution by randomly flipping the signs of the allele
frequency changes per-genomic window. This destroys the systematic covariances
created by linked selection and creates a sampling distribution of the
covariances spuriously created by neutral genetic drift while preserving the
complex dependencies between adjacent loci created by linkage disequilibrium.
This empirical neutral null distribution is conservative in the sense that the
variances of the covariances are wider than expected under drift alone as they
include the effect of selection on the allele frequency change within a
time-interval, just not between time-intervals. We see (Figure
\ref{fig:figure-3} A and B) that windowed temporal covariances between close
timepoints are skewed positive (a heavy right tail), while between more distant
timepoints these windowed temporal covariances tend to shift to become more
negative (a heavy left tail).  We quantified the degree to which the left and
right tails are inflated compared to the null distribution as a function of
time, and see excesses in both tails in Figure \ref{fig:figure-3} C. This
finding is also robust to sign-permuting allele frequency changes on a
chromosome-level, the longest extent that gametic linkage disequilibria can
extend (Supplementary Figure \ref{suppfig:barghi-tailprobs-seqid}). We see a
striking pattern that the windowed covariances not only decay towards zero, but
in fact become negative through time, consistent with many regions in the
genome having had a reversed fitness effect at later timepoints.

@@

Finally we used forward-in-time simulations to explore the conditions under
which temporal and convergent correlations arise. We show a subset of our
results for a model of Gaussian stabilizing selection (GSS) on a
phenotype  \gc{DO WE actually need to say Gaussian here? Or just in
  methods. }
where directional selection is induced by a sudden shift in the optimum
phenotype of varying magnitudes (Figure \ref{fig:figure-4}A).  We find that
positive temporal covariances are produced by such selection (Figure
\ref{fig:figure-4}B), and that these positive temporal covariances can compound
together to generate a large proportion of allele frequency change being due to
selection (i.e. large G) over the relatively short time periods similar to our
analyzed selection datasets span (Figure \ref{fig:figure-4}C). The magnitude of
these covariances increases with the strength of selection, i.e. the variance
in fitness, such that stronger selection generates larger proportions of allele
frequency change.  We find a similar picture of stronger convergent selection
pressures generating larger convergence correlations \gc{(Figure {fig:figure-4}D)}.

These results are often relatively insensitive to the number of loci underlying
the trait, suggesting that they may be capturing highly polygenic signals.
Indeed if only a small number of loci influence the trait, the $G(t)$
trajectories are typically much more stochastic, such that replicates do not
reliably generate temporal covariances (compare Figure \ref{fig:figure-1}B to
Supplementary Material Figure \gc{XXX}). \gc{This again suggests that
  the genome-wide linked selection response we see in the empirical data is highly
polygenic. Using our simulations we find that sampling only every 10
generations does indeed mean that our estimates of $G(t)$ are an
underestimate of the contribution of linked selection as they cannot include the covariance between closely
spaced generations (see XXXX).
}

Additionally, we explored other selection schemes. We find that the long term
dynamics of the covariances under \gc{directional} truncation selection, which generates
substantial epistasis, are richer than we see under GSS and multiplicative
selection (Supplementary Material Figure XXX).  We also conducted simulations
of purifying selection alone (i.e. background selection) and find that this can
also generate positive temporal covariances (Figure XXX), thus it is unlikely
that all of the allele frequency change we see is the result of the novel
selection pressure the populations are exposed to, and some of this selection
may be ongoing purifying selection. Only in the case of the Longshanks
experiment, does the control line mean that we can quantify selection that is
due to the novel selection pressure.

While none of our experiments have selected the populations in divergent
directions, in our simulations we find that such selection can generate
negative convergent correlations \gc{(Figure {fig:figure-4}D)}. This suggests that selection experiments
combining multiple replicates, control lines, as well as divergent selection
pressures might be quite informative for unpacking the contribution of
particular selection pressures to genome-wide allele frequency change.

@@

Our analyses reveal that a sizable proportion of allele frequency change in
these populations is due to the (indirect) action of selection. Capitalizing on
replicated designs, we characterized the extent to which convergent selection
pressures lead to parallel changes in allele frequencies across replicate
populations, and found that a reasonable proportion of the response is shared
across short timescales. These likely represent substantial under-estimates of
the contribution of linked selection because the studies we have reanalyzed do
not sequence the population each generation, preventing us from including the
effects of stronger correlations between adjacent generations.  Furthermore,
our estimation methods are intentionally conservative, for example they exclude
the contribution of selection that does not persist across generations and
selection that reverses sign; thus they can be seen as a strong lower bound of
the effects of selection.

@@

%By using an exponential fitness function (corresponding to
%multiplicative selection across sites), \vb{we ensure there is no build up of
%LD beyond what is initially present due to selection and epistasis}
%\parencite{Turelli1990-kd,Bulmer1980-zo}; this serves as the simplest
%directional selection model of a trait to understand the effects of genetic
%architecture on the statistics we have used in the paper.

@@

\subsubsection{Sampling in Temporal Blocks}
